Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-08-13 review_article.tex
Fix quotes

@@

A choropleth map is used to show the spatial relationship, or
variability, of measurements. Created as a storage device which
preserves locations for geographically ordered data, with usage dating
back to the 1800s (Berry, Morrill, and Tobler,
\protect\hyperlink{ref-GOINO}{n.d.}). They are constructed by drawing
the given geographic or political boundaries, and filling the shapes
with colours to represent values of a measured variable (Tufte
\protect\hyperlink{ref-EI}{1990}). Early versions used symbols or
patterns instead of colour. Bell et al.
(\protect\hyperlink{ref-CPISACA}{2006}) discusses the use of choropleth
maps for visualising cancer data, and Walter
(\protect\hyperlink{ref-DMAHP}{2001}) gives an overview of the
development of the use of choropleth maps for displaying disease data. A
choropleth map is a true map of the topology, constructed for visual
inspection of spatial patterns across a familiar geographic form. Figure
1 shows a choropleth map where each state of the United States of
America is coloured by the average annual rate of new cases of lung and
bronchus cancer from years 2012 to 2016.

\begin{figure}
\centering
\includegraphics{figures/ggchoro.png}
\caption{``A choropleth map of the United States of America.''}
\end{figure}

Utilising the familiar state boundaries can make a map intuitive to read
(Brewster and Subramanian \protect\hyperlink{ref-CIBMUK}{2010}), and
allow viewers to visually infer the spatial relationships in the data,
i.e.~how cancer rate differs across the states. The familiarity of the
geography is a worthy consideration when presenting results of spatial
analysis. Just as geographers are no longer the only creators of maps,
Bell et al. (\protect\hyperlink{ref-CPISACA}{2006}) suggests the
audiences of spatial health data analysis have extended beyond
researchers to the public, policymakers and the media. However, while
the areas are recognisable shapes, they are often politically driven
boundaries with individual areas being of non-uniform size, containing
different population densities and subject to change over time. The
different population and geographical sizes of administrative areas can
attract attention to the shades of the unpopulated but large areas
(Tufte \protect\hyperlink{ref-EI}{1990}). Choropleths can inhibit visual
inference when presenting human related statistics as the display may
draw attention from the `potentially more important results in the more
populous communities' (Exeter \protect\hyperlink{ref-SE}{2017}).

In epidemiology, choropleths are often used as a tool to study the
spatial distribution of cancer incidence and mortality.

Almost 100 years of cancer mapping in the United States and the United
Kingdom has increased the effectiveness in the presentation of unbiased
rates. The increasing development and use of disease maps can be
attributed to the availability of geographic information system software
(Exeter \protect\hyperlink{ref-SE}{2017}). The data collection methods
of cancer mortality rates across regions, and administrative control
within regions lends itself to choropleth visualisation. Mortality rates
are now often presented as relative rates of risk across the population,
and age adjusted to correct for the the higher prevalence of cancers in
older populations. Howe (\protect\hyperlink{ref-HEDP}{1989}) describes
Stocks development of the standardised mortality ratios through the
1930s. The choropleth maps used at this time presented levels of cancer
via hatchings on a black and white scale.

Mortality rates are now often presented as relative rates of risk across
the population, and age adjusted. Howe
(\protect\hyperlink{ref-HEDP}{1989}) describes Stocks' development of
the standardised mortality ratios through the 1930s. The choropleth maps
used presented levels of cancer via hatchings on a black and white
scale.

Examples of atlases: Kraak 1998, Kraak and Ormeling 2011 Bertin 1967

\hypertarget{public-face}{%
\subsection{Public face}\label{public-face}}

A grey literature review conducted by (ref) identified 33 publicly
available cancer atlases published on the internet between 01/01/2010 to
11/11/2015. Reflecting the high use of chloropleth maps founds in the
literature, outlined in the section above, all the public facing maps
identified in this review were also chloropleth maps.

\hypertarget{publishers}{%
\subsubsection*{Publishers}\label{publishers}}
\addcontentsline{toc}{subsubsection}{Publishers}

The majority of atlases were published by non-commercial organisations,
including not-for-profits (NFPs), government, research organisations,
advocacy groups or a partnership between an NFP \& government. Only one
map was published by a commercial entity (Maps of Cancer Mortality Rates
in Spain), in this case a media organisation El Pais\footnote{\url{https://elpais.com/}}.

\hypertarget{reported-measures}{%
\subsubsection*{Reported measures}\label{reported-measures}}
\addcontentsline{toc}{subsubsection}{Reported measures}

The majority of maps identified in this study reported age adjusted
rates of either incidence, mortality or survival. The list below
summarises these measures, and provides a definition of each measure.

\hypertarget{geographical-coverage}{%
\subsubsection*{Geographical coverage}\label{geographical-coverage}}
\addcontentsline{toc}{subsubsection}{Geographical coverage}

Identified cancer atlases covered geographies from all around the world:
4 were global, 3 from Australia (AUS), 11 from the United States (US), 2
from Canada (CAN), 7 from the United Kingdom (UK), 2 from Spain, 1 from
Switzerland, 1 from Germany, 1 from Norway, and 1 covering the European
Union. Not all maps had a national focus and 10 covered a region or
state rather than an entire nation. The states or counties/regions
covered were South Australia (AUS), Queensland (AUS), Ontario (CAN),
Valencia (Spain), Pennsylvania county Massachusetts (US), New Hampshire
(US), Cape Cod (US), Missouri (US), Florida (US), New York State (US)
and Arizona (US).

\newpage

\begin{longtable}[]{@{}ll@{}}
\caption{(\#tab:measures) Measures used to report cancer
statistics}\tabularnewline
\toprule
\begin{minipage}[b]{0.28\columnwidth}\raggedright
Measure\strut
\end{minipage} & \begin{minipage}[b]{0.66\columnwidth}\raggedright
Details\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{0.28\columnwidth}\raggedright
Measure\strut
\end{minipage} & \begin{minipage}[b]{0.66\columnwidth}\raggedright
Details\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.28\columnwidth}\raggedright
1. IR (Incidence Ratio)\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
\((IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}\),\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Cancer incidence rate in region \(i\) over the average cancer incidence
rate for the total region\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
2. SIR (Standardised Incidence Ratio)\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
IR standardised by age structure in each region \(i\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
3. RER\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
\(RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
(Relative Excess Risk)\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Represents the estimate of cancer related mortality within five years of
diagnosis\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Also referred to as `excess hazard ratio'\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
4. Age Adjusted Relative Risk\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
RR standardised by age structure in each region \(i\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
5. Rate per 100,000\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Cancer incidence per 100,000 population\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
6. Age Adjusted Rate per 100,000\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
\#5 standardised by age structure or region\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
7. New cancer cases per 100,000\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Specific methods could not be found\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
8. Count\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Crude cancer counts\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
9. Below or above Expected\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Alternative expression of the SIR\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{statistical-uncertainty}{%
\subsubsection{Statistical uncertainty}\label{statistical-uncertainty}}

Cancer atlases were considered to report uncertainty to the non-expert
user if they included a measure of statistical uncertainty either within
or alongside the map. Maps that only reported this information within
the supplementary material were not considered to have directly
attempted to report uncertainty.

The review did not reveal any novel uncertainty visualisation approaches
or visualisations. Maps used standard and well known measures including
credible intervals and standard deviation, statistical significance, box
plots and distributions. These maps ranged from static pdfs or
infographics to interactive online resources. The interactivity of the
more modern maps enabled uncertainty information to be incorporated
without cluttering the screen, such as in a tool tip feature.

@@

{\centering \includegraphics[width=0.8\linewidth]{figures/Alberta_CI_viz}

@@

\caption{Example of CI visualisation for uncertainty representation in cancer mapping (1/3). Source: Alberta Health IHDA Geographic. (2012) }\label{fig:ci-viz2}
\end{figure}

{\centering \includegraphics[width=0.55\linewidth]{figures/Pennsylvania_CI_viz}

\caption{Example of CI visualisation for uncertainty representation in cancer mapping (2/3). Source: Pensylvania Cancer Atlas }\label{fig:ci-viz3}
\end{figure}

Close to half of the atlases identified (42\%, n=14) included some
measure of uncertainty. The most common measure used to represent
uncertainty were credible or confidence intervals (CIs). CIs were either
visualised by including their bounds in a scatterplot or graph of
estimates vs region (see Figures @ref(fig:ci-viz2)\footnote{Alberta
  Health IHDA Geographic. (2012). Age-Standardised Incidence Rate of
  COPD, 2011. Retrieved from:
  \url{http://www.health.alberta.ca/health-info/IHDA-geographic/COPD/incidence-agestandard/atlas.html?epik=0GJSpE_IW34lx}},
and @ref(fig:ci-viz3)\footnote{Centres for Disease Control and
  Prevention (CDC). (n.d). United States Cancer Statistics: An
  Interactive Cancer Atlas (InCA). Retrieved
  from:\url{https://nccd.cdc.gov/DCPC_INCA/}} positioned next to the
map, or reported numerically through the CI upper and lower bounds
listed in a data table (see Figure 3.4\footnote{Pennsylvania Cancer
  Atlas. (n.d). Retrieved from:
  \url{https://www.geovista.psu.edu/grants/CDC/?epik=0lJSpE_IW34lx}}).
Of those that visualised the CIs, 30\% (n=10) embedded the visualisation
within a tool tip function which visualised the CI when the mouse
hovered over the relevant area (see Figure 3.5\footnote{International
  Agency for Research on Cancer. (2017). Atlas of Cancer Mortality in
  the European Union and European Economic Area 1993-1997, Annex 4 -
  Cancer mortality maps by site. Retrieved from:
  \url{http://www.iarc.fr/en/publications/pdfs-online/epi/sp159/}}).

{\centering \includegraphics[width=0.6\linewidth]{figures/CDC_CI_viz_tooltip}

\caption{Example of CI visualisation for uncertainty representation in cancer mapping (3/3). Source: Centres for Disease Control and Prevention (CDC). United States Cancer Statistics: An Interactive Cancer Atlas (InCA) }\label{fig:ci-tooltip}
\end{figure}

{\centering \includegraphics[width=0.5\linewidth]{figures/uncert_table}

\caption{Example of an interactive data table with CI upper and lower bounds used in cancer mapping. Source: Pennsylvania Cancer Atlas}\label{fig:ci-table}
\end{figure}

@@

{\centering \includegraphics[width=0.4\linewidth]{figures/Boxplot2}

\caption{Example of boxplot used in cancer mapping. Source: Source: Pensylvania Cancer Atlas }\label{fig:boxplot-example}
\end{figure}

@@

\hypertarget{cartograms}{%
\subsection{Cartograms}\label{cartograms}}

Choropleths may be considered true topological maps, however, if the
land mass presented covers enough of the globe, there must be a
transformation or distortion to display in 2D. The amount of distortion
is related to the distance displayed Tobler
(\protect\hyperlink{ref-GAMP}{1963}). The world projections reflect the
frequent distortions seen from altering perspectives. Choropleth maps
will always be distorted if they cover enough of the globe, as will
photographs of the globe from space. Choropleth creation requires
choosing a map projection that shows a favourable distortion of the
geography for presenting the set of spatial information. Diagrams that
do not specify a projection can be considered to have some unknown
projection.

@@

Waldo Tobler (\protect\hyperlink{ref-TVSSS}{2012}) explores many
graphical techniques, and suggests there are particular methods for
particular purposes. Cartograms provide an alternative visualisation
method for statistical and geographical information. The key difference
between a choropleth and a cartogram is the desirable augmentation of
the size, shape or distance of geographical areas (Dorling
\protect\hyperlink{ref-ACTUC}{2011}). Cartograms may be seen as an
extension of map transformations and projections. The favourable
distortion is proportional to a value other other than actual earth size
area (Olson \protect\hyperlink{ref-NAC}{1976}).

Using choropleth maps for population characteristics requires graphic
distortions when the population concerned varies greatly in density
(Griffin \protect\hyperlink{ref-CTTMB}{1980}). When implementing a
distortion of the geographical shape according to population, an area
cartogram (Olson \protect\hyperlink{ref-NAC}{1976}), or iso-demographic
map is the result. When visualising population statistics Dorling
(\protect\hyperlink{ref-ACTUC}{2011}) considers this equitable
representation design `more socially just, giving due attention to all
members of the population and reducing the visual impact of large areas
with small populations (Walter \protect\hyperlink{ref-DMAHP}{2001}).
Griffin (\protect\hyperlink{ref-CTTMB}{1980}) suggest that spatial
socio-economic data is best presented on a cartogram for urban areas.
Howe (\protect\hyperlink{ref-HEDP}{1989}) agrees that 'cancer occurs in
people, not in geographical areas' and the map bases of population
reflect this and avoid allocating `undue prominence' to rural areas.
Jahan et al. (\protect\hyperlink{ref-MTMSIH}{2018}) encourage the use of
cartograms to highlight small areas and uncover local-level
inequalities. Using health estimates from large areas can prevent
drawing attention to the inequalities.

The area on the map space can also be used represent a value. There have
been many algorithms presented, Nusrat and Kobourov
(\protect\hyperlink{ref-SAIC}{2016}) provided a framework to investigate
implementations and the ``statistical accuracy, geographical accuracy,
and topological accuracy''. The spatial transformation of map regions
relative to the data emphasises the data distribution instead of land
size (Kocmoud and House \protect\hyperlink{ref-CBATCC}{1998}).

Cancer statistics and atlases

The creation of cartograms was largely in the hands of cartographers.
Dorling (\protect\hyperlink{ref-ACTUC}{2011}) discusses early approaches
including John Hunter and Jonathan Young (1968) and Durham's wooden tile
method, Skoda and Robertson's (1972) steel ball bearing approach and
Tobler's (1973) computer programs.

@@

`Firstly, that no such transformation has a unique solution, leaving the
subjective choice in the hands of the individual map designer.'

Design goals: - to maintain as much as possible of the intricate detail
- to maximize the simplicity of the transformation - maintenance of
directional accuracy - attempt to influence and direct the map user's
attention,

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  that the size of the electoral subdivisions should be made
  proportional to their respective electoral populations,
\item
  that the contiguity relationships between the DCU's should be
  maintained,
\item
  that true directions should be retained from the centre of the
  transformation,
\item
  that the main shape characteristics of the DCU's should be preserved,
  and
\item
  that the map should possess the overall shape of the built-up area.
\end{enumerate}

@@

A contiguous cartogram allows the map space to highlight the
distribution of the variable. These cartograms maintain connectivity of
the map regions while areas are resized according to a statistic, this
often must occur at the expense of the shape (Kocmoud and House
\protect\hyperlink{ref-CBATCC}{1998},@NAC, @TAAM). From a computer
graphics perspective, Min Ouyang and Revesz
(\protect\hyperlink{ref-ACA}{2000}) believe it is a problem of `map
deformation' to account for the value assigned to each area, they
provide three methods for creating value-by-area cartograms. Examples
include Tobler's Pseudo-Cartogram Method, Dorling's Cellular Automaton
Method (\protect\hyperlink{ref-ACTUC}{2011}), Radial Expansion Method of
Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and
Tikunov's Line Integral Method, Constraint-Based Method (Kocmoud and
House) (\protect\hyperlink{ref-CBATCC}{1998}).

>>>>

Maps have been adopted to present geospatial statistics for centuries.
Maps connect data and statistics to the geographical representation of
areas that are generally familiar to the audience. However, it is not
enough for areas on maps to be recognisable, if information regarding
the distribution of the statistic cannot be gleaned. In situations where
the people are of interest, not the land they live on, it is reasonable
to explore views that enhance the communication of the cancer
statistics. Identifying and explaining spatial structures, patterns, and
processes involves considering the individuals in communities and
organising communities into representable units (Moore and Carpenter
\protect\hyperlink{ref-SAMGIS}{1999}). This has spurred innovations over
the previous centuries to enhance the shapes and maps presented to
effectively communicate cancer outcomes, and health outcomes more
broadly. Cancer statistics directly represent and relate to the people
living within individual geographical areas.

The visualisation methods used to present cancer statistics will depend
on the intended message and users of the map. Transforming statistics to
visualisations considers individual observations or observation points
aggregated into communities and geographical units. Visualising diseases
on maps is often the first step in exploratory spatial data analysis and
effectively helps in the formulation of hypotheses (Jahan et al.
\protect\hyperlink{ref-MTMSIH}{2018}). Disease maps help to present
geographic patterns that may have been overlooked in a table, obscuring
the geospatially related statistics (Moore and Carpenter
\protect\hyperlink{ref-SAMGIS}{1999}). By providing a visual
representation of cancer outcomes, geographic patterns of disease are
able to be identified and effectively addressed.

@@

A choropleth map is used to display the spatial characteritics of a
relationship, or variability, of measurements. As an alternative storage
device to a table, it preserves locations for geographically ordered
data, with usage dating back to the 1800s (Berry, Morrill, and Tobler,
\protect\hyperlink{ref-GOINO}{n.d.}). A choropleth is constructed by
drawing the geographic or political boundaries, and filling the shapes
with colours to represent values of a measured variable (Tufte
\protect\hyperlink{ref-EI}{1990}). Early versions of choropleth maps
used symbols or patterns instead of colour. Bell et al.
(\protect\hyperlink{ref-CPISACA}{2006}) discusses their use in
visualising cancer data, and Walter
(\protect\hyperlink{ref-DMAHP}{2001}) gives an overview of the
development of these maps for displaying disease data. Just as
geographers are no longer the only creators of maps, Bell et al.
(\protect\hyperlink{ref-CPISACA}{2006}) suggests the audiences of
spatial health data analysis have extended beyond researchers to the
public, policymakers and the media. A choropleth map is a true map of
the topology, constructed for visual inspection of spatial patterns
across a familiar geographic form. Figure 1 shows a choropleth map where
each state of the United States of America is coloured by the average
annual rate of new cases of lung and bronchus cancer from years 2012 to
2016.

@@

{\centering \includegraphics[width=0.8\linewidth]{figures/ggchoro}

@@

\caption{A choropleth map of the United States of America.}\label{fig:choro}
\end{figure}

Utilising the familiar state boundaries can make a map intuitive to read
(Brewster and Subramanian \protect\hyperlink{ref-CIBMUK}{2010}), and
allow viewers to visually infer the spatial relationships in the data,
i.e.~how cancer rate differs across the states. The familiarity of the
geography is a worthy consideration when presenting results of spatial
analysis. Just as geographers are no longer the only creators of maps,
Bell et al. (\protect\hyperlink{ref-CPISACA}{2006}) suggests the
audiences of spatial health data analysis have extended beyond
researchers to the public, policymakers and the media. While the areas
are recognisable shapes, they are often politically driven boundaries
with individual areas being of non-uniform size, containing different
population densities and subject to change over time. The different
population and geographical sizes of administrative areas can attract
attention to the shades of the unpopulated but large areas (Tufte
\protect\hyperlink{ref-EI}{1990}). Choropleths can inhibit visual
inference when presenting human related statistics as the display may
draw attention from the `potentially more important results in the more
populous communities' (Exeter \protect\hyperlink{ref-SE}{2017}).

Choropleth maps can be useful devices for communicating information to
public on a familiar map base. In epidemiology, choropleths are often
used as a tool to study the spatial distribution of cancer incidence and
mortality. A cancer atlas is a choropleth map, or collection of maps,
representing cancer incidence and mortality for a country, or group of
countries. d'Onofrio et al. (\protect\hyperlink{ref-MACM}{2016}) provide
the definition and reports that cancer atlases began with Haviland's
maps in 1875. The data collection methods of cancer mortality rates
across regions, and the administrative control within regions lends
itself to choropleth visualisation. The increasing development and use
of disease maps can be attributed to the availability of geographic
information system software (Exeter \protect\hyperlink{ref-SE}{2017}).
Early work in US cancer atlases can be attributed to Burbank ???, and UK
cancer atlases to Howe 1963 ???. The choropleth maps presented levels
via hatchings or dots on a black and white scale. These atlases were key
to developing hypotheses regarding areas with unusually high rates,
geographic correlations, work related exposures, and high risk diets
d'Onofrio et al. (\protect\hyperlink{ref-MACM}{2016}).

Almost 100 years of cancer mapping in the United States and the United
Kingdom has seen increased effectiveness in the presentation cancer
statistics. Mortality rates are now often presented as relative rates of
risk across the population, and age adjusted to correct for the the
higher prevalence of cancers in older populations. Howe
(\protect\hyperlink{ref-HEDP}{1989}) describes Stock's development of
the standardised mortality ratios through the 1930s. Table 1 presents
summarises the measures presented in published cancer atlases, and
provides a definition of each measure.

Table 1: Measures used to report cancer statistics

\begin{longtable}[]{@{}ll@{}}
\toprule
\begin{minipage}[b]{0.28\columnwidth}\raggedright
Measure\strut
\end{minipage} & \begin{minipage}[b]{0.66\columnwidth}\raggedright
Details\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.28\columnwidth}\raggedright
1. IR (Incidence Ratio)\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
\((IR)_i=\frac{(Incidence\ Rate)_i}{Average\ Incidence\ Rate}\),\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Cancer incidence rate in region \(i\) over the average cancer incidence
rate for the total region\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
2. SIR (Standardised Incidence Ratio)\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
IR standardised by age structure in each region \(i\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
3. RER\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
\(RER = \frac{(Cancer\ related\ mortality)_i}{Average\ cancer\ related\ mortality}\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
(Relative Excess Risk)\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Represents the estimate of cancer related mortality within five years of
diagnosis\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Also referred to as `excess hazard ratio'\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
4. Age Adjusted Relative Risk\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
RR standardised by age structure in each region \(i\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
5. Rate per 100,000\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Cancer incidence per 100,000 population\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
6. Age Adjusted Rate per 100,000\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
\#5 standardised by age structure or region\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
7. New cancer cases per 100,000\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Specific methods could not be found\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
8. Count\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Crude cancer counts\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.28\columnwidth}\raggedright
9. Below or above Expected\strut
\end{minipage} & \begin{minipage}[t]{0.66\columnwidth}\raggedright
Alternative expression of the SIR\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{supporting-material-in-cancer-atlases}{%
\subsubsection{Supporting material in cancer
atlases}\label{supporting-material-in-cancer-atlases}}

A map communicates quickly and draws attention to prominent geographies,
but an atlas is often supplemented with supporting statistics.
Cruickshank's (1947) as cited by Walter
(\protect\hyperlink{ref-DMAHP}{2001}), discusses using visuals as
`formal statistical assessment of the spatial pattern'. However, when
presenting cancer atlases, d'Onofrio et al.
(\protect\hyperlink{ref-MACM}{2016}) believes the intuition derived must
be `validated by rigorous statistical analyses.'

These additional statistics often include a measure of the statistical
uncertainty of the values of the statistics presented in a choropleth.
In the review of atlases in the public domain, atlases were considered
to report uncertainty to the non-expert user if they included a measure
of statistical uncertainty either within or alongside the map. Maps that
only reported this information within the supplementary material were
not considered to have directly attempted to report uncertainty. The
maps considered used standard and well known measures including credible
intervals and standard deviation, statistical significance, box plots
and distributions. These maps ranged from static pdfs or infographics,
to interactive online resources.

The interactivity of the more modern maps enabled supplementary
information to be incorporated without cluttering the screen, such as in
a tool tip feature.

\hypertarget{publicly-available-atlases-using-choropleth-methods}{%
\subsection{Publicly available atlases using choropleth
methods}\label{publicly-available-atlases-using-choropleth-methods}}

Cancer maps are effective communication tools for a general or
non-expert audience. They are commonly used in the public domain to
communicate results of sophisticated statistical analyses. The heavy use
of chloropleth maps within the research literature is reflected in the
types of maps that are found in the public domain. A grey literature
review conducted by (ref) identified 33 cancer atlases published on the
internet between January 2010 and November 2015. These chloropleth maps
were mostly published by non-commercial organisations, including
not-for-profits (NFPs), government, research organisations, advocacy
groups or a partnership between an NFP \& government. Only one map was
published by a commercial entity.

The cancer atlases covered geographies from all around the world, four
were global atlases. Most focussed on single nations, the United States
was considered by eleven atlases, the United Kingdom by seven, followed
by three of Australia, two of Canada, and one of each from Switzerland,
Germany, Norway. One atlas covered the European Union. Not all maps had
a national focus and ten covered a region or state rather than an entire
nation. The states or counties/regions covered were South Australia
(AUS), Queensland (AUS), Ontario (CAN), Valencia (Spain), Pennsylvania
county Massachusetts (US), New Hampshire (US), Cape Cod (US), Missouri
(US), Florida (US), New York State (US) and Arizona (US).

\hypertarget{examples-of-supporting-statistics-in-public-facing-atlases}{%
\subsubsection{Examples of supporting statistics in public facing
atlases}\label{examples-of-supporting-statistics-in-public-facing-atlases}}

Close to half of the atlases identified (42\%, n=14) included some
measure of uncertainty. The most common measure used to represent
uncertainty were credible or confidence intervals (CIs).

@@

@@

\hypertarget{cartograms-as-an-alternative-display}{%
\subsection{Cartograms as an alternative
display}\label{cartograms-as-an-alternative-display}}

A cartogram alters the map base with the intention of improving the
presentation of the statistic of interest. For a single variable of
interest, each map area is changed to emphasise the distribution by
representing the corresponding value, in comparison to the value of the
other areas (Dougenik, Chrisman, and Niemeyer
\protect\hyperlink{ref-ACCAC}{1985}).

Choropleths may be considered true topological maps, however, if the
land mass displayed covers enough of the globe, there must be a
transformation or distortion to display the land in 2D. The amount of
distortion is related to the distance covered by the landmass displayed
Tobler (\protect\hyperlink{ref-GAMP}{1963}). World map projections
reflect the frequent perspectives used to view the earth. Choropleth
maps will always be distorted if they cover enough of the globe, just as
photographs of the globe from space. Choropleth creation requires
choosing a map projection that shows a favourable distortion of the
geography for presenting the set of spatial information. Diagrams that
do not specify a projection can be considered to have some unknown
projection.

If the statistic presented on the map base relies on physical distance
and is influenced by the topology there is no transformation needed
beyond choosing a projection. The purposeful distortion of the map space
is beneficial when a uniform density of the map base is desired. When a
map base is transformed according to population density, population
becomes a uniformly distributed background for the statistic presented
(Berry, Morrill, and Tobler, \protect\hyperlink{ref-GOINO}{n.d.}).

@@

If the statistic presented on the map base relies on physical distance
and is influenced by the topology there is no transformation needed
beyond choosing a projection. The purposeful distortion of the map space
is beneficial when a uniform density of the map base is desired, Dorling
(\protect\hyperlink{ref-ACTUC}{2011}) suggests `population distribution
is often extremely uneven in former British colonies', this makes the
distortion necessary. When a map base is transformed according to
population density, population becomes a uniformly distributed
background for the statistic presented (Berry, Morrill, and Tobler,
\protect\hyperlink{ref-GOINO}{n.d.}). Using choropleth maps for
population characteristics requires graphic distortions when the
population concerned varies greatly in density (Griffin
\protect\hyperlink{ref-CTTMB}{1980}). When implementing a distortion of
the geographical shape according to population, an area cartogram (Olson
\protect\hyperlink{ref-NAC}{1976}), population-by-area cartograms
(Levison and Haddon Jr \protect\hyperlink{ref-TAAM}{1965}), or
iso-demographic map is the result.

Cartograms provide an alternative visualisation method for statistical
and geographical information. The key difference between a choropleth
and a cartogram is the desirable augmentation of the size, shape or
distance of geographical areas (Dorling
\protect\hyperlink{ref-ACTUC}{2011}). Cartograms may be seen as an
extension of map transformations and projections. The favourable
distortion is proportional to a value other other than actual earth size
area (Olson \protect\hyperlink{ref-NAC}{1976}). A disadvantage of the
conventional map is that sparsely populated rural areas may be
emphasized, whereas the areas representing cities are very small, making
interpretation of spatial patterns very difficult. The distortion of a
cartogram prevents the density obscuring the spatial patterns (Levison
and Haddon Jr \protect\hyperlink{ref-TAAM}{1965}). The spatial
transformation of map regions relative to the data emphasises the data
distribution instead of land size (Kocmoud and House
\protect\hyperlink{ref-CBATCC}{1998}). When visualising population
statistics Dorling (\protect\hyperlink{ref-ACTUC}{2011}) considers this
equitable representation design `more socially just', or honest (Dent
\protect\hyperlink{ref-NISCC}{1972}), giving due attention to all
members of the population and reducing the visual impact of large areas
with small populations (Walter \protect\hyperlink{ref-DMAHP}{2001}).
Griffin (\protect\hyperlink{ref-CTTMB}{1980}) suggest that spatial
socio-economic data is best presented on a cartogram for urban areas.
Howe (\protect\hyperlink{ref-HEDP}{1989}) agrees that `cancer occurs in
people, not in geographical areas' and the map bases of population
reflect this and avoid allocating `undue prominence' to rural areas.
Jahan et al. (\protect\hyperlink{ref-MTMSIH}{2018}) encourage the use of
cartograms to highlight small areas and uncover local-level
inequalities.

The creation of cartograms was largely in the hands of professional
cartographers (Kraak \protect\hyperlink{ref-CD}{2017}). Dorling
(\protect\hyperlink{ref-ACTUC}{2011}) discusses early approaches
including John Hunter and Jonathan Young (1968) and Durham's wooden tile
method, Skoda and Robertson's (1972) steel ball bearing approach and
Tobler's (1973) computer programs. Howe
(\protect\hyperlink{ref-HEDP}{1989}) discusses the impact of electronic
computer-assisted techniques. Geographical information systems allowed
map users, and researchers to implement their own cartograms, but these
systems are utilised depending on `the effectiveness, efficiency, and
satisfaction of the map products (Nielsen 1994)'(Kraak
\protect\hyperlink{ref-CD}{2017}).

There have been many algorithms presented, Nusrat and Kobourov
(\protect\hyperlink{ref-SAIC}{2016}) provided a framework to investigate
implementations and the ``statistical accuracy, geographical accuracy,
and topological accuracy''.

@@

@@

A contiguous cartogram maintains connectivity of the map regions while
areas are resized according to a statistic. This transformation often
occurs at the expense of the shape of areas (Kocmoud and House
\protect\hyperlink{ref-CBATCC}{1998}, @NAC, @TAAM). From a computer
graphics perspective, Min Ouyang and Revesz
(\protect\hyperlink{ref-ACA}{2000}) believe it is a problem of `map
deformation' to account for the value assigned to each area, they
provide three methods for creating value-by-area cartograms. Examples
include Tobler's Pseudo-Cartogram Method, Dorling's Cellular Automaton
Method (\protect\hyperlink{ref-ACTUC}{2011}), Radial Expansion Method of
Selvin et al., Rubber Sheet Method of Dougenik et al., Gusein-Zade and
Tikunov's Line Integral Method, Constraint-Based Method (Kocmoud and
House) (\protect\hyperlink{ref-CBATCC}{1998}).

An intentional goal when creating the 1966 Census population cartogram
for Canada was to maintain contiguity, while attempting to keep the
actual shape of places. The end result was a `very accurate
isodemo-graphic map of Canada'. This intentional design goal coincided
with the rising interest in urban geography.
